---
title: "The statistical value chain and data validation"
author: "Mark van der Loo and Edwin de Jonge"
date: "uRos2019 Tutorial Session, Bucharest"
output: 
  beamer_presentation:
    fig_caption: FALSE
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This tutorial


`https://github.com/data-cleaning/uRos2019_tutorial`



# The Statistical Value Chain


## Statistical Value Chain

\begin{center}
\includegraphics[height=3cm]{fig/valuechain.pdf}
\end{center}


### Notes

- This part only pertains to the data processing stage. Collection, design,
dissemination is not included.
- The fixed points are well-defined statistical products.



## The SVC: Remarks

- Actual data processing is not necessarily linear accross the chain
- In production architectures a more flexible model is often used where
the definition of interfaces between processing steps play a crucial role.
The chain shown here is a general example covering most steps in some way.


## Data validation

### Definition (ESS handbook on validation)

_Data validation is an activity in which it is verified whether or not a
combination of values is a member of a set of acceptable value combinations._

### Validation rules

The set of acceptable values combinations are defined by 
_validation rules_, e.g. `IF age <= 14 THEN has_job == "no"`.

### Observe

_validation rules define, to large extend, the products in the SVC_


## `validate`: _data validation infrastructure for R_

### A domain-specific language for rule definition

Define _any_ check on your data, using the _full power_ of the R 
language.

### Rules as first-class citizens

- CRUD operations (create, read, update, delete)
- Summarize, plot, investigate rules
- Rich metadata

### Validate data

- Confront data with rules
- CRUD on results, summarize, plot
- Export to ESS standard reporting format (upcoming)

## Assignment 1

Try the following code.

```{r, results='hide'}
library(validate)
library(magrittr)
data(retailers)
head(retailers)
retailers %>%
  check_that(turnover + other.rev == total.rev
             , turnover > 0, other.rev > 0 ) %>%
  summary()
```

## Assignment 1


\scriptsize{}
```{r}
library(validate)
library(magrittr)
data(retailers) 
retailers %>%
  check_that(turnover + other.rev == total.rev
             , turnover > 0, other.rev > 0 ) %>%
  summary()
```
\normalfont{}


## Data validation with `validate`

\scriptsize{}
```{r}
library(validate)
data(retailers)
head(retailers,3)[3:7]
```
\normalfont{}


## Data validation with `validate`


```{r,results='hide'}
rules <- validator(
   turnover >= 0
  , other.rev >= 0
  , turnover + other.rev == total.rev
)

out <- confront(retailers, rules)
summary(out)
```


## Assignment 2

1. Adapt the previous exercise so you use `validator`.
2. Use `confront` for validation and store the results in a variable called `out`.
3. Try `plot(out)`.
4. Try `as.data.frame(out)` (use `View` to inspect the result)


## Plotting output

\scriptsize{}
```{r,fig.height=5}
plot(out)
```
\normalfont{}


## The validate package

![](fig/overview.pdf)


## Reading rules from file

```{r, eval=FALSE}
### myrulez.txt

# some basic checks
staff >= 0
turnover >= 0 
other.rev >= 0
# account balance checks
turnover + other.rev == total.rev
# other commom sense stuff
if (staff >= 1) staff.costs >= 1
```

```{r,eval=FALSE}
rulez <- validator(.file="myrulez.txt")
```


## Assignment 3

1. Create a new textfile
2. Define 10 rules for the `retailers` dataset
3. Read the rules (`validator(.file="your file")`)
4. `confront` rules with data
5. Summarize and plot the results.
6. Use `as.data.frame` and `View` to convert and display the results.
7. Make a `plot` of the `validator` object.

# A few extra's (if we have time)

## Domain Specific Language

### Validation _Domain Specific Language_ (DSL)

Any R statement resulting in a `logical`.

### Examples

```{r, eval=FALSE}
# Range checks
has_job %in% c('yes','no')
turnover >= 0
# Multivariate checks
abs(profit) <= 0.6 * turnover
# Multi-row checks
mean(profit) > 10
# Logical implications
if (staff > 0) staff.costs > 0
```

## Validation DSL

### Comparisons

`>, >=,==, <=, <, %in%`

### Boolean operations

`!, all(), any(), &, &&, |, ||, if () else`

### Text search

`grepl`


### Functional dependencies (Armstrong)

`city + zipcode ~ streetname`

### Refer to the dataset with `.`

`nrow(.) == 40`, `"turnover" %in% names(.)`

## Transient assignments (macros) using `:=`

### Example 1

$$
\max\left(\frac{x}{x^*},\frac{x^*}{x}\right) \leq 10
$$

```{r, eval=FALSE}
med := median(turnover,na.rm=TRUE)
hb := pmax(turnover/med, med/turnover, na.rm=TRUE)
hb <= 10
```

### Example 2

```{r,eval=FALSE}
beta_2 := coefficients(lm(turnover ~ profit))[2]
beta_2 >= 0
```


## Variable groups

### Many variables, same rule

```{r, eval=FALSE}
G := var_group(staff, turnover, other.rev, total.costs)
G >= 0
```

## Error handling

```{r}
out <- check_that(women, hite > 0, weight>0)
out
errors(out)
```




