---
title: "Finding and handling data errors"
subtitle: "errorlocate" 
author: "Edwin de Jonge and Mark van der Loo"
date: "uRos2019 Tutorial Session, Bucharest"
output:
  beamer_presentation:
    keep_tex: no
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
library(errorlocate)
library(magrittr)
```

## Error localization

Data validation and error localization answer different questions.

### Data validation

Which errors are there? 

### Error localization

Where do I need to make changes to fix the errors?


## Example

### Ruleset

```{r,eval=FALSE}
age >= 0
age <= 120
if (drivers_licence == TRUE) age >= 18
```

### Data

```{r, echo=FALSE}
knitr::kable(data.frame(age=10,drivers_licence=TRUE))
```

### Question:

Which field or fields would you change?



## Error localization

### Definition

Error localization is a procedure that points out fields in a
data set that can be altered or imputed in such a way that
all validation rules can be satisfied.


## Example

### Ruleset

```{r, eval=FALSE}
if (married == TRUE ) age >= 16
if (attends == "kindergarten") age <= 6
```

### Data

```{r, echo=FALSE, eval=TRUE}
knitr::kable(data.frame(age=3, married=TRUE, attends="kindergarten"))
```


### Question

Which field or fields would you change?

## Principle of Fellegi and Holt

###

Find the minimal (weighted) number of fields to adjust such that
all rules, including implied rules, can be satisfied.

\hfill{}\scriptsize{IP Fellegi and D Holt, JASA \textbf{71} 353  17--35 (1976).}

### Note

This should be used as a last resort, when no further information on 
the location of errors is available.


## Implied rules?

### 

```{r, eval=FALSE}
turnover - total.cost == profit
               profit <= 0.6 * turnover
```

This implies (substituting profit):

```{r, eval=FALSE}
           total.cost >= 0.4 * turnover
```

We need to take into account such _essentially new_ rules (edits) ---unstated 
relations between variables that can be derived from the explicitly defined rules.


<!--
## Finding errors in records.

\begin{center}
\includegraphics[width=\textwidth]{../validation/fig/valuechain}
\end{center}

## Validation rules?

As we saw package `validate` allows to:

- formulate explicit data rule that data must conform to:

```{r}
library(validate)
rules <- validator(
  age >= 0, 
  age < 150,
  if (driver_license == TRUE) age >= 16
)
```


## Explicit validation rules:

- Give a clear overview what the data must conform to.
- Can be used to fix/correct data!
- Find error, and when found correct it.

### Note:

- Manual editing is error prone, not reproducible and not feasible for large data sets.
- Large rule sets have (very) complex behavior, e.g. entangled rules: adjusting one value may 
invalidate other rules.

## 

### `validate` tells us:

- which **records** are **invalid**
- which data **rules** are **violated**.

### `errorlocate` tells us:

- which field/**variable**(s) should be fixed to make a record valid.

## Data example

```{r}
rules <- validator(
  age >= 0, 
  age < 150,
  if (driver_license == TRUE) age >= 16
)

# invalid data
car_owners <- data.frame(age = 160, driver_license = TRUE)
```

Question: _Which variable (likely) is  incorrect? Why?_

## Data example

```{r}
rules <- validator(
  age >= 0, 
  age < 150,
  if (driver_license == TRUE) age >= 16
)

# invalid data
car_owners <- data.frame(age = 160, driver_license = TRUE)
```

**Clearly `age` is incorrect, because it violates the second constraint.**

## Data example 2

```{r}
rules <- validator(
  age >= 0, 
  age < 150,
  if (driver_license == TRUE) age >= 16
)

car_owners <- data.frame(age = 10, driver_license = TRUE)
```

Question: _Which variable (likely) is  incorrect? Why?_

-->

## Data example 2

```{r, eval=FALSE}
age >= 0, 
age < 150,
if (driver_license == TRUE) age >= 16
```

```{r, echo=FALSE}
knitr::kable(data.frame(age = 10, driver_license = TRUE))
```

### Question

Which variable would you adjust?

**It depends on the quality of `age` and `driver_license`. We can add more weight to `age` if we think that variable has better quality.**

## Assignment: 

```{r}
rules <- validator(
  if (married == TRUE) age >= 16,
  if (attends == "kindergarten") age <= 6
) 

persons <- data.frame( age     = 3
                     , married = TRUE
                     , attends = "kindergarten"
                     )
```

a) check with `validate` which rules are violated.
b) What should be changed to this record to "correct" it? Why? 

## Feligi Holt (FH) formalism:

> Find the minimal (weighted) number of variables that cause the invalidation of the data rules.

### 

Makes sense if no further knowledge on the error mechanism is available! 

R package `errorlocate` (second generation of `editrules`) implements FH.

## Feligi Holt (FH) formalism:

But there are exceptions...

  - In balance sheets, swapping variables (2 edits) sometimes makes more sense then adjusting one value (1 edit). (see R package:`deducorrect`).

  - In some data,  spreading a surplus or shortage on a variable over many variables is sensible. (see R package: `rspa`).

## `errorlocate`

- R-package that implements FH.
- Is extensible (you can plug in your own detection stuff)
- provides:
  
    - `locate_errors`
    - `replace_errors`
    - R5 classes to add your own stuff.

## `errorlocate::locate_errors`

```{r, eval=TRUE}
locate_errors( data.frame( age  = 3
                  , married = TRUE
                  , attends = "kindergarten"
                  )
     , validator( if (married == TRUE) age >= 16
                , if (attends == "kindergarten") age <= 6
                )
     )
```

## `errorlocate::locate_errors`

```{r, eval=TRUE}
locate_errors( data.frame( age  = 3
                  , married = TRUE
                  , attends = "kindergarten"
                  )
     , validator( if (married == TRUE) age >= 16
                , if (attends == "kindergarten") age <= 6
                )
     )$errors
```

## Assignment (small examples)

a) Find the error in this record with `locate_errors`: 

```{r, echo=FALSE}
knitr::kable(data.frame( age = 26, married = TRUE, attends= "kindergarten"))
```

b) Find the error with `locate_errors`: 

```{r, echo=FALSE}
knitr::kable(data.frame( age = 15, married = TRUE
          , attends= "kindergarten"))
```


## Removing errors

- Detecting errors is very useful, but then what? 
- Fixing philosophy is: 
  - Find erroneuous values.
  - Remove them (i.e. make them `NA`).
  - Impute them with sensible values.
  
### Note
We could also remove erroneous records completely, but often this result in _over-deletion_ and introduces a _bias_.

## `errorlocate::replace_errors`

- Locates errors and replaces them with `NA`.

```{r, eval=TRUE}
replace_errors( 
    data.frame( age     = 3
              , married = TRUE
              , attends = "kindergarten"
              )
  , validator( if (married == TRUE) age >= 16
             , if (attends == "kindergarten") age <= 6
             )
)
```

<!--

## Assignment

a) Use the data set `retailers` from package `validate`.
b) Use validate to find out which records are `faulty` using the rule set

```{r, echo = TRUE, eval=FALSE}
rules <- validator(
  to_pos = turnover >= 0
  , or_pos = other.rev >= 0
  , balance = turnover + other.rev == total.rev)

```

c) use `locate_errors` to find some errors.
d) use `replace_errors` to "fix" the data set.

## 

```{r, eval = TRUE}

data(retailers, package="validate")
retailers <- retailers[c("other.rev", "total.rev", "turnover")]
rules <- validator(
  to_pos = turnover >= 0
  , or_pos = other.rev >= 0
  , balance = turnover + other.rev == total.rev)
confront(retailers, rules)

errors <- locate_errors(retailers, rules)$errors
row_contains_error <- apply(errors, 1, any)
w <- which(row_contains_error)
errors[w,]
```

##

```{r, eval=TRUE}
retailers[w,c("other.rev", "total.rev", "turnover")]
replace_errors(retailers[w,], rules)
```

--> 

## Internal workings:

`errorlocate`:

- translates error localization problem into a **mixed integer problem**, which
is solved with `lpsolveAPI`.

- contains a small framework for implementing your own error localization algorithms.


## Pipe friendly

The `replace_errors` function is pipe friendly:

```{r, eval=FALSE}
rules <- validator(age < 150)

data_noerrors <- 
  data.frame(age=160, driver_license = TRUE) %>% 
  replace_errors(rules)

errors_removed(data_noerrors) # contains errors removed
```

